A Transactional Memory System for Parallel Python

### 2014-07-29 talk/dls2014/paper/paper.tex
re-add mentioning non-x86; explain better why JIT-STM matters less at the moment; make JIT-STM result summary sound less pessimistic

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in Section~\ref{subsec:jit-benchs}. The STM-JIT
integration is currently still incomplete and not tested much. The
JIT-less interpreter provides a much more consistent environment for
the STM system, so we remove some unknown variables by disabling it.

@@

Still, STM rarely beats CPython's single-thread performance. However, for
programs that need concurrency in CPython and that use threads to
achieve this, it also makes sense to look at the overhead induced by
the GIL on multiple threads. From this perspective, the STM
implementation beats CPython's performance in all but two benchmarks.

@@

Overall PyPy needs the JIT  for its performance to be
competitive with CPython's. It would be interesting to see how using
our STM system in CPython would turn out, but an in-depth evaluation is
beyond the scope of this paper. On its own, our system scales well,
so we expect to also see this trend in the presence of a
JIT in the future.

@@

To obtain an overall performance that is competitive with the
best-performing Python systems (Jython, PyPy), integration of the
STM-based approach with a JIT compiler is necessary. Once this
integration matures, the approach outlined here serves not only as a
simple GIL replacement but also provides a way forward towards
a parallel programming model for Python.

>>>>

On non-x86 architectures, most simple memory accesses could still be
done efficiently if the supported addressing modes allow for the
addition of an offset stored in some register (e.g.\ ARM). For more
complicated accesses (e.g.\ array indexing) or if the CPU does not
support such an addressing mode, one extra addition may be required.

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in Section~\ref{subsec:jit-benchs}. The STM-JIT
integration is currently still incomplete and not tested much. The
JIT-less interpreter provides a much more consistent environment for
the STM system, so we remove some unknown variables by disabling it.
Furthermore, we think the interpreter-STM evaluation is more relevant
at this stage as the results can be more directly applied to other
similar interpreters. PyPy's JIT~\cite{cfbolz09}, however, is quite
unique as it is a JIT tracing the interpreter instead of the
interpreted language itself. This demands its own thorough
evaluation, which is out of the scope of this paper.

@@

Still, STM rarely beats CPython's \emph{single-thread} performance. However, for
programs that need concurrency in CPython and that use threads to
achieve this, it also makes sense to look at the overhead induced by
the GIL on multiple threads. From this perspective, the STM
implementation beats CPython's performance in all but two benchmarks.

@@

Overall, PyPy without STM is around $2\times$ slower than CPython.
Enabling its JIT allows it to outperform CPython by a huge margin.
We see the same kind of speedup on PyPy with STM when enabling the
JIT. This means that STM does not generally inhibit the JIT from
optimising the programs execution, which is already a very important
result on its own. We also see that for some
benchmarks, STM is already able to give additional speedups
compared to just the JIT-induced acceleration. This looks very
promising and investigating when this is not the case is the next
logical step.

@@

To generally outperform the best-performing Python systems (Jython,
PyPy), integration of the STM-based approach with a JIT compiler is
necessary. Our early results of this integration suggest that there is
no inherent incompatibility between STM and PyPy's JIT.  Once the
implementation matures, the approach outlined here serves not only as a
simple GIL replacement but also provides a way forward towards a
parallel programming model for Python.
